Transcriptomic responses to diet quality and viral infection in Apis mellifera

### 2018-11-06 Manuscript/Text/manuscript.tex
Adam additional changes

\author[
   addressref={aff1},                   % id's of addresses, e.g. {aff1,aff2}
   email={lrutter@iastate.edu}   % email address
]{\inits{LR}\fnm{Lindsay} \snm{Rutter}}
\author[
   addressref={aff6},
   email={bbonning@ufl.edu}
]{\inits{BB}\fnm{Bryony C.} \snm{Bonning}}
\author[
   addressref={aff2},
   email={dicook@monash.edu}
]{\inits{DC}\fnm{Dianne} \snm{Cook}}
\author[
   addressref={aff3,aff4},
   email={amytoth@iastate.edu}
]{\inits{AT}\fnm{Amy L.} \snm{Toth}}
\author[
   addressref={aff5},
   corref={aff5},
   email={adolezal@illinois.edu}
]{\inits{AD}\fnm{Adam G.} \snm{Dolezal}}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                          %%
%% Enter the authors' addresses here        %%
%%                                          %%
%% Repeat \address commands as much as      %%
%% required.                                %%
%%                                          %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\address[id=aff1]{%                           % unique id
  \orgname{Bioinformatics and Computational Biology Program, Iowa State University},
  \city{Ames},   
  \state{IA}
  \postcode{50011},
  \cny{USA}
}
\address[id=aff2]{%
  \orgname{Econometrics and Business Statistics, Monash University},
  \city{Clayton},   
  \state{VIC}
  \postcode{3800},
  \cny{Australia}
}

\address[id=aff3]{%
  \orgname{Department of Entomology, Iowa State University},
  \city{Ames},   
  \state{IA}
  \postcode{50011},
  \cny{USA}
}

\address[id=aff4]{%
  \orgname{Department of Ecology, Evolution, and Organismal Biology, Iowa State University},
  \city{Ames},   
  \state{IA}
  \postcode{50011},
  \cny{USA}
}

\address[id=aff5]{%
  \orgname{Department of Entomology, University of Illinois at Urbana-Champaign},
  \city{Urbana},   
  \state{IL}
  \postcode{61801},
  \cny{USA}
}

\address[id=aff6]{%
  \orgname{Department of Entomology and Nematology, University of Florida},
  \city{Gainesville},   
  \state{FL}
  \postcode{32611},
  \cny{USA}
}

@@

Commercially managed honey bees have undergone unusually large declines in the United States and parts of Europe over the past decade \cite{ccd1, ccd2, ccd3}, with annual mortality rates exceeding what beekeepers consider sustainable \cite{ccd5, ccd6}. More than 70 percent of major global food crops (including fruits, vegetables, and nuts) at least benefit from pollination, and yearly insect pollination services are valued worldwide at \$175 billion \cite{ccd7}. As honey bees are largely considered to be the leading pollinator of numerous crops, their marked loss has considerable implications for agricultural sustainability \cite{ccd4}.

Honey bee declines have been associated with several factors, including pesticide use, parasites, pathogens, habitat loss, and poor nutrition \cite{factors, factors2}. Researchers generally agree that these stressors do not act in isolation; instead, they appear to influence the large-scale loss of honey bees in an interactive fashion as the environment changes \cite{interacting}. Nutrition and viral infection are two broad factors that pose heightened dangers to honey bee health in response to recent environmental changes.

@@

We also compare the main effect of IAPV exposure in our dataset to that obtained in a previous study conducted by Galbraith and colleagues \cite{galbraith}. While our study examines honey bees from naturally mated colonies, the Galbraith study examined honey bees from single-drone colonies. As a consequence, the honey bees in our study will be on average 25\% genetically identical, whereas honey bees from the Galbraith study will be on average 75\% genetically identical \cite{sisters}. We note that the difference between these studies may be even greater than this as we used naturally mated honey bees from 15 different colonies. We should therefore expect that the Galbraith study may generate data with higher signal:to:noise ratios than our data due to lower genetic variation between its replicates. At the same time, our honey bees will be more likely to display the health benefits gained from increased genotypic variance within colonies, including decreased parasitic load \cite{multParasite}, increased tolerance to environmental changes \cite{divHyp2}, and increased colony performance \cite{geneticDiverse, geneticDiverse2}. Given that honey bees are naturally very polyandrous \cite{patriline}, our naturally mated honey bees may also reflect more realistic environmental and genetic simulations. Taken together, each study provides a different point of value: Our study likely presents less artificial data while the Galbraith data likely presents less messy data. We wish to explore how the gene expression effects of IAPV inoculation compare between these two studies that used such different experimental designs. To achieve this objective, we use visualization techniques to assess the signal:to:noise ratio between these two datasets, and differential gene expression (DEG) analyses to determine any significantly overlapping genes of interest between these two datasets. As RNA-sequencing data can be biased \cite{biased1, biased2, biased3}, this comparison allowed us to characterize how repeatable and robust our RNA-sequencing results were in comparison to previous studies. It also allowed us to shine light on how experimental designs that control genetic variability to different extents might affect the resulting gene expression data in honey bees. We suggest that in-depth data visualization approaches can be useful for cross-study comparisons and validation of noisy RNA-sequencing data in the future.

@@

Given the noisy nature of our data, and our desire to hone in on genes with real expression differences, we compared our data to the Galbraith study \cite{galbraith}, which also examined bees response to IAPV infection. In contrast to our study, Galbraith et al. identified a large number of virus responsive transcripts, and generally had less noise in their data (Figure 3 and Additional files 2A, 5, 6, 7, 8, 9, and 10). To identify the most consistent virus-responsive genes from our study, we looked for overlap in the DEGs associated with virus infection on both experiments. We found a large, statistically significant (p-value $<$ 2.2e-16) overlap, with 26/38 (68\%) of virus-responsive DEGs from our study also showing response to virus infection in Galbraith et al. (Figure 6). This result gives us confidence that, although noisy, we were able to uncover reliable, replicable gene expression responses to virus infection with our data.

@@

There were several limitations in this study that could be improved upon in future studies. For instance, our comparison between the Galbraith data (single-drone colonies) and our data (naturally mated colonies) was limited by numerous extraneous variables between these studies. In addition to different molecular pipelines and bioinformatic preprocessing pipelines used between these studies, the Galbraith study focused on one-day old worker honey bees that were fed sugar and artificial pollen diet, whereas our study focused on adult worker honey bees that were fed bee-collected monofloral diets. Furthermore, the Galbraith data used eviscerated abdomens with attached fat bodies and only considered symptomatic honey bees for their infected treatment group, whereas we used whole bodies and considered both asymptomatic and symptomatic honey bees for our infected treatment group. There are also differences in the hours post inoculation and possible differences in the inoculation amount between the studies. Further differences between the studies can be found in their corresponding published methods sections \cite{adamInt, galbraith}. The different factors between these two studies may be critical because particular antiviral factors in honey bees are linked to specific viruses, specific developmental stages, the analyzed tissue, the route of inoculation, and the time (post-inoculation) during which the study was performed. This was clearly demonstrated when comparing honey bee responses to two related iflaviruses with very different infection dynamics, sacbrood bee virus (SBV) vs. deformed wing virus (DWV) \cite{Ryabov}. Authors observed differences in induction of defensin and hymenoptaecin immune-related genes, and suggested the results reflect adaptations to the different routes of transmission \cite{Ryabov}.

Moreover, our comparative visualization assessment between these two datasets was also somewhat limited because the virus effect in the Galbraith study used three replicates for each level, whereas the virus effect in our study used twelve replicates for each level that were actually further subdivided into six replicates for each diet level. Hence the apparent reduction in noise observed in the Galbraith data compared to our data in the PCA plots, parallel coordinate plots, scatterplot matrices, and litre plots may be an inadvertent product of the smaller number of replicates used and the lack of a secondary treatment group rather than solely the reduction in genetic variability through the single-drone colony design itself. With this in mind, while our current efforts may be a starting point, future studies can shed more light on signal:to:noise and differential expression differences between naturally mated colony designs and single-drone colony designs by controlling for extraneous factors more strictly than what we were able to do in the current line of work.

@@

We used honey bees from naturally mated colonies, whereas Galbraith et al. \cite{galbraith} used honey bees from single-drone colonies. In light of this, we should expect the Galbraith et al. dataset to contain lower genetic variation between its replicates and higher signal:to:noise ratios than our dataset. We use visualization techniques to assess the signal:to:noise ratio between these two datasets, and differential gene expression (DEG) analyses to determine any significantly overlapping genes of interest between these two datasets.

@@

\section*{Author's contributions}
  LR performed the bioinformatic and statistical analyses, produced the figures and tables, and drafted the manuscript. BB conceptualized the study and critically revised the manuscript. AD contributed to experimental design, carried out the laboratory experiments, and processed samples for virus titers and RNA-seq. 
DC advised on statistical analyses and visualization.
    %All authors read and approved the final manuscript.

>>>>

\author[
   addressref={aff1}, 
   email={lrutter@iastate.edu} 
]{\inits{LR}\fnm{Lindsay} \snm{Rutter}}
\author[
   addressref={aff2}, 
   email={jcarrillo@cicese.mx} 
]{\inits{LR}\fnm{Jimena} \snm{Carrillo-Tripp}}
\author[
   addressref={aff3},
   email={bbonning@ufl.edu}
]{\inits{BB}\fnm{Bryony C.} \snm{Bonning}}
\author[
   addressref={aff4},
   email={dicook@monash.edu}
]{\inits{DC}\fnm{Dianne} \snm{Cook}}
\author[
   addressref={aff5,aff6},
   email={amytoth@iastate.edu}
]{\inits{AT}\fnm{Amy L.} \snm{Toth}}
\author[
   addressref={aff7},
   corref={aff7},
   email={adolezal@illinois.edu}
]{\inits{AD}\fnm{Adam G.} \snm{Dolezal}}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                          %%
%% Enter the authors' addresses here        %%
%%                                          %%
%% Repeat \address commands as much as      %%
%% required.                                %%
%%                                          %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\address[id=aff1]{%                           % unique id
  \orgname{Bioinformatics and Computational Biology Program, Iowa State University},
  \city{Ames},   
  \state{IA}
  \postcode{50011},
  \cny{USA}
}

\address[id=aff2]{%
  \orgname{Department of Microbiology, Center for Scientific Research and Higher Education of Ensenada},
  \city{Ensenada},   
  \state{Baja California}
  \postcode{22860},
  \cny{Mexico}
}

\address[id=aff3]{%
  \orgname{Department of Entomology and Nematology, University of Florida},
  \city{Gainesville},   
  \state{FL}
  \postcode{32611},
  \cny{USA}
}

\address[id=aff4]{%
  \orgname{Econometrics and Business Statistics, Monash University},
  \city{Clayton},   
  \state{VIC}
  \postcode{3800},
  \cny{Australia}
}

\address[id=aff5]{%
  \orgname{Department of Entomology, Iowa State University},
  \city{Ames},   
  \state{IA}
  \postcode{50011},
  \cny{USA}
}

\address[id=aff6]{%
  \orgname{Department of Ecology, Evolution, and Organismal Biology, Iowa State University},
  \city{Ames},   
  \state{IA}
  \postcode{50011},
  \cny{USA}
}

\address[id=aff7]{%
  \orgname{Department of Entomology, University of Illinois at Urbana-Champaign},
  \city{Urbana},   
  \state{IL}
  \postcode{61801},
  \cny{USA}
}

@@

Managed honey bees have undergone unusually large declines in the United States and parts of Europe over the past decade \cite{ccd1, ccd2, ccd3}, with annual mortality rates exceeding what beekeepers consider sustainable \cite{ccd5, ccd6}. More than 70 percent of major global food crops (including fruits, vegetables, and nuts) at least benefit from pollination, and yearly insect pollination services are valued worldwide at \$175 billion \cite{ccd7}. As honey bees are largely considered to be the leading pollinator of numerous crops, their marked loss has considerable implications for agricultural sustainability \cite{ccd4}.

Honey bee declines have been associated with several factors, including pesticide use, parasites, pathogens, habitat loss, and poor nutrition \cite{factors, factors2}. Researchers generally agree that these stressors do not act in isolation; instead, they appear to influence the large-scale loss of honey bees in an interactive fashion as the environment changes \cite{interacting}. Nutrition and viral infection are two factors that pose heightened dangers to honey bee health in response to recent environmental changes. Interactions between nutrition and viral infection may create feedbacks that impact bee health through several mechanisms \cite{adamInt, AdamTothReview}.

@@

We also compare the main effect of IAPV exposure in our dataset to that obtained in a previous study conducted by Galbraith and colleagues \cite{galbraith}. While our study examines honey bees derived from naturally-mated queens, the Galbraith study examined honey bees derived from single-drone inseminated queens. As a consequence, the honey bees in our study will be on average 25\% genetically identical, whereas honey bees from the Galbraith study will be on average 75\% genetically identical \cite{sisters}. We note that the difference between these studies may be even greater than this as we used honey bees from 15 different colonies, i.e. from 15 different, naturally-mated queens. We should therefore expect that the Galbraith study may generate data with higher signal:to:noise ratios than our data due to lower genetic variation between its replicates. At the same time, our honey bees will be more likely to display the health benefits gained from increased genotypic variance within colonies, including decreased parasitic load \cite{multParasite}, increased tolerance to environmental changes \cite{divHyp2}, and increased colony performance \cite{geneticDiverse, geneticDiverse2}. Given that honey bees are naturally very polyandrous \cite{patriline}, our naturally-mated honey bees may also reflect more realistic environmental and genetic conditions. Taken together, each study provides a different point of value: Our study likely presents less artificial data while the Galbraith data likely presents less messy data. We therefore wish to explore how the gene expression effects of IAPV inoculation compare between these two studies using different experimental designs. To achieve this objective, we use visualization techniques to assess the signal:to:noise ratio between these two datasets, and differential gene expression (DEG) analyses to determine any significantly overlapping genes of interest between these two datasets. As RNA-sequencing data can be biased \cite{biased1, biased2, biased3}, this comparison allowed us to characterize how repeatable and robust our RNA-sequencing results were in comparison to previous studies. It also allowed us to shine light on how experimental designs that control genetic variability to different extents might affect the resulting gene expression data in honey bees. We suggest that in-depth data visualization approaches can be useful for cross-study comparisons and validation of noisy RNA-sequencing data in the future.

@@

Given the noisy nature of our data, and our desire to home in on genes with real expression differences, we compared our data to the Galbraith study \cite{galbraith}, which also examined bees response to IAPV infection. In contrast to our study, Galbraith et al. identified a large number of virus responsive transcripts, and generally had less noise in their data (Figure 3 and Additional files 2A, 5, 6, 7, 8, 9, and 10). To identify the most consistent virus-responsive genes from our study, we looked for overlap in the DEGs associated with virus infection on both experiments. We found a large, statistically significant (p-value $<$ 2.2e-16) overlap, with 26/38 (68\%) of virus-responsive DEGs from our study also showing response to virus infection in Galbraith et al. (Figure 6). This result gives us confidence that, although noisy, we were able to uncover reliable, replicable gene expression responses to virus infection with our data.

@@

There were several limitations in this study that could be improved upon in future studies. For instance, our comparison between the Galbraith data (single-drone colonies) and our data (naturally-mated colonies) was limited by numerous extraneous variables between these studies. In addition to different molecular pipelines and bioinformatic preprocessing pipelines used between these studies, the Galbraith study focused on worker honey bees that were fed sugar and artificial pollen diets, whereas our study focused on worker honey bees that were fed bee-collected monofloral diets. Furthermore, the Galbraith data used eviscerated abdomens with attached fat bodies and only considered symptomatic honey bees for their infected treatment group, whereas we used whole bodies and considered both asymptomatic and symptomatic honey bees for our infected treatment group. There are also differences in the hours post inoculation and possible differences in the inoculation amount between the studies. Further differences between the studies can be found in their corresponding published methods sections \cite{adamInt, galbraith}. The different factors between these two studies may be critical because particular antiviral factors in honey bees are linked to specific viruses, specific developmental stages, the analyzed tissue, the route of inoculation, and the time (post-inoculation) during which the study was performed. This was clearly demonstrated when comparing honey bee responses to two related iflaviruses with very different infection dynamics, sacbrood bee virus (SBV) vs. deformed wing virus (DWV) \cite{Ryabov}. Authors observed differences in induction of defensin and hymenoptaecin immune-related genes, and suggested the results reflect adaptations to the different routes of transmission \cite{Ryabov}.

Moreover, our comparative visualization assessment between these two datasets was also somewhat limited because the virus effect in the Galbraith study used three replicates for each level, whereas the virus effect in our study used twelve replicates for each level that were actually further subdivided into six replicates for each diet level. Hence the apparent reduction in noise observed in the Galbraith data compared to our data in the PCA plots, parallel coordinate plots, scatterplot matrices, and litre plots may be an inadvertent product of the smaller number of replicates used and the lack of a secondary treatment group rather than solely the reduction in genetic variability through the single-drone colony design itself. With this in mind, while our current efforts may be a starting point, future studies can shed more light on signal:to:noise and differential expression differences between naturally-mated colony designs and single-drone colony designs by controlling for extraneous factors more strictly than what we were able to do in the current line of work.

@@

We used honey bees from naturally-mated colonies, whereas Galbraith et al. \cite{galbraith} used honey bees from single-drone colonies. In light of this, we should expect the Galbraith et al. dataset to contain lower genetic variation between its replicates and higher signal:to:noise ratios than our dataset. We use visualization techniques to assess the signal:to:noise ratio between these two datasets, and differential gene expression (DEG) analyses to determine any significantly overlapping genes of interest between these two datasets.

@@

\section*{Author's contributions}
  LR performed the bioinformatic and statistical analyses, produced the figures and tables, and drafted the manuscript. BB conceptualized the study and critically revised the manuscript. AD contributed to experimental design, carried out the laboratory experiments, and processed samples for virus titers and RNA-seq. JCT contributed to experimental design and laboratory experiments. DC advised on statistical analyses and visualization.
    %All authors read and approved the final manuscript.
